---
title: "HIV testing uptake among women in South Africa"
author: "Dickman Pangaume Gareta"
date: "`r date()`"
format: 
  html:
    toc: true
    #code-fold: show
    #code-tools: true
    df-print: paged
    embed-resources: true
  pdf:
   toc: true
  docx: 
   toc: true   
theme: cayman
editor_options: inlinew
---

## Introduction

HIV is a public health concern worldwide and efforts are underway to eliminate the virus by 2030. However, knowing one's HIV status is crucial as it provides an entry point into HIV care, treatment and lifelong support. This is especially important in regions with the highest burden of HIV such as the Sub-Saharan Africa. There is documented evidence that HIV incidence is high among women especially those in the older age groups. Therefore, understanding HIV status among women is key in preventing the acquisition of the disease or linking to HIV care and treatment.

The objective of the project is to evaluate the uptake of HIV testing by social demographic factors among women aged between 15 and 49 years in **South Africa**. The data source for this analysis is the *2016 South Africa Demographic and Health Surveys(DHS)*. More info about the survey can be found [here](https://dhsprogram.com/data/Guide-to-DHS-Statistics/).



```{r include = F}

#| include = false
# Clear working directory

rm(list = ls())
gc()

```



```{r include = F}
# Load packages

if(!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, here,haven, lubridate,unibeCols,janitor,usethis )
```



```{r}
#| label: data
#| include: false
# Load data
sa_dhs <- haven::read_dta(here("data/ZAIR71FL.DTA"),
  col_select = c(
    caseid,
    v008, # Date of interview (CMC
    v011, #  Date of birth (CMC)
    v012, # Respondent's current age
    v013, # Age in 5-year groups
    v015, # Result of individual interview
    v025, # Type of place of residence
    v102, # Type of place of residence
    v106, # Highest educational level
    v107, #  Highest year of education
    v133, # Education in single years
    v149, # Educational attainment
    v190, #  Wealth index combined
    v190a, # Wealth index for urban/rural
    v437, # Respondent's weight in kilograms (1 decimal)
    v438, # Respondent's height in centimeters (1 decimal)
    v445, # Body Mass Index
    v454, # Currently pregnant
    v463a, # Smokes cigarettes
    v463aa, # Frequency smokes cigarettes
    v171a, # Use of internet
    v171b, #  Frequency of using internet last month
    v120, # Household has: radio
    v121, #  Household has: television
    v781, # Ever been tested for HIV
    v535, # Ever been married or in union
    v536, #  Recent sexual activity
    v531, # Age at first sex (imputed)
    v761, #  Condom used during last sex with most recent
    v763a, # Had any STI in last 12 months,
    v453  # Hemoglobin level (g/dl - 1 decimal) 
  )
)

```



```{r include = F}

# Rename variables

sa_dhs_df <- sa_dhs |>
  rename(
    idno = caseid,
    dob = v008,
    dbo1 = v011,
    age = v012,
    age_cat = v013,
    interview_result = v015,
    residence = v025,
    educ_level = v106,
    educ_level_yr = v107,
    educ_yr = v133,
    educ_attain = v149,
    wealth_index_comb = v190,
    wealth_index = v190a,
    weight = v437,
    height = v438,
    bmi = v445,
    preg_current = v454,
    smoker = v463a,
    smoke_freq = v463aa,
    internet_use = v171a,
    internet_use_month = v171b,
    hh_radio = v120,
    hh_tv = v121,
    hiv_test_ever = v781,
    married_ever = v535,
    recent_sex = v536,
    age_first_sex = v531,
    condom_use_lsex = v761,
    had_sti_12m = v763a,
    hemoglobin = v453  
  )


tempfile(pattern = "_rs_rdf_", fileext = ".rdf")

haven::as_factor(sa_dhs_df) # Converting the labelled data to R factors
summary(sa_dhs_df$hemoglobin)

```



```{r include = F}

# Exploratory data analysis

head(sa_dhs_df)
dim(sa_dhs_df)
n_distinct(sa_dhs_df)


```


```{r include = F}
sa_dhs_df1 <- sa_dhs_df |>
  mutate(
    hiv_test_ever_cat = case_when(
      hiv_test_ever == 1 ~ "Tested before",
      hiv_test_ever == 0 ~ "Never tested before",
      TRUE ~ NA
    ),
    hiv_test_ever_cat = as.factor(hiv_test_ever_cat) # Change the outcome variable to a factor variable
  ) |> select(idno,hiv_test_ever, hiv_test_ever_cat, age, bmi, everything())

```





```{r echo = F}

# New variables

sa_dhs_df2 <- sa_dhs_df1 |>
  mutate(
    wealth_index_cat = case_when(
      wealth_index == 1 ~ "Poorest",
      wealth_index == 2 ~ "Poorer",
      wealth_index == 3 ~ "Middle",
      wealth_index == 4 ~ "Richer",
      wealth_index == 5 ~ "Richest",
      TRUE ~ NA
    ),
    .before = wealth_index,
    
    # Adding levels to wealth indices
    
    wealth_index_cat = factor(
      wealth_index_cat,
      levels = c("Poorest",
                 "Poorer",
                 "Middle",
                 "Richer",
                 "Richest")
    ),
    # education level
    educ_level_cat = case_when(
      educ_level == 0 ~ "No education",
      educ_level == 1  ~ "Primary",
      educ_level == 2 ~ "Secondary",
      educ_level == 3 ~ "Higher"
    ),
    educ_level_cat = factor(
      educ_level_cat,
      levels = c("No education",
                 "Primary",
                 "Secondary",
                 "Higher")
    ),
    age = as.integer(age),
    hemoglobin = as.numeric(hemoglobin / 10),
    age_first_sex = as.integer(age_first_sex),
    
    # residence type
    
    residence_type = case_when(residence == 1 ~ "Rural",
                               residence == 2 ~ "Urban",
                               TRUE ~ "Missing"),
    weight_kg = (as.numeric(weight)) / 10,
    #  Adding 1 decimal to the weight in Kg
    height_m = (as.numeric(height) / 10) / 100,
    # Adding 1 decimal to the height in cm and convert the height in cm to meters
    bmi = (as.numeric(bmi) / 100),
    
    married_ever = case_when(
      married_ever == 0 ~ "No",
      married_ever == 1 ~ "Yes",
      married_ever == 2 ~ "Lived with a man",
      TRUE ~ "Missing"
    ),
    
    married_ever = factor(married_ever, levels = c("No",
                                                   "Yes",
                                                   "Lived with a man",
                                                   "Missing")),
    
    preg_current = case_when(preg_current == 0 ~ "No",
                             preg_current == 1 ~ "Yes",
                             TRUE ~ "Missing"),
    
    smoker = case_when(smoker == 0 ~ "No",
                       smoker == 1 ~ "Yes",
                       TRUE ~ ""),
    recent_sex = case_when(
      recent_sex == 0 ~ "Never had sex",
      recent_sex == 1 ~ "Active",
      recent_sex == 2 ~ "Not active",
      recent_sex == 3 ~ "Not active",
      TRUE ~ "Missing"
    ),
    recent_sex = factor(recent_sex, levels = c("Never had sex",
                                               "Active",
                                               "Not active")),
    condom_use_lsex = case_when(
      condom_use_lsex == 0 ~ "No",
      condom_use_lsex == 1 ~ "Yes",
      TRUE ~ "Missing"),
    condom_use_lsex = factor(condom_use_lsex, levels = c("No",
                                                         "Yes",
                                                         "Missing")),
    had_sti_12m = case_when(had_sti_12m == 0 ~ "No",
                            had_sti_12m == 1 ~ "Yes",
                            TRUE ~ "Missing"),
    had_sti_12m = factor(had_sti_12m, levels = c("No",
                                                 "Yes",
                                                 "Missing")),
    age_cat = case_when(
      age_cat == 1 ~ "15 - 19",
      age_cat == 2 ~ "20 - 24",
      age_cat == 3 ~ "25 - 29",
      age_cat == 4 ~ "30 - 34",
      age_cat == 5 ~ "35 - 39",
      age_cat == 6 ~ "40 - 44",
      age_cat == 7 ~ "45 - 49",
      age_cat == 8 ~ "50+",
      TRUE ~ "Missing"
    ),
    age_cat = factor(
      age_cat,
      levels = c(
        "15 - 19",
        "20 - 24",
        "25 - 29",
        "30 - 34",
        "35 - 39",
        "40 - 44",
        "45 - 49",
        "50+",
        "Missing"
      )
    ),
    internet_use = case_when(
      internet_use == 0 ~ "No",
      internet_use >= 1 &  internet_use <= 3 ~ "Yes",
      TRUE ~ "Missing"
    ),
    hh_radio = case_when(hh_radio == 0 ~ "No",
                         hh_radio == 1 ~ "Yes",
                         TRUE ~ "Missing"),
    
    hh_tv = case_when(hh_tv == 0 ~ "No",
                      hh_tv == 1 ~ "Yes",
                      TRUE ~ "Missing"),
    hh_radio = factor(hh_radio, levels = c("No",
                                           "Yes",
                                           "Missing")),
    hh_tv = factor(hh_tv, levels = c("No",
                                     "Yes",
                                     "Missing"))
  )



```

# Background characteristics

```{r include = F}

# Summary statistics for continuous variables

sa_descrip <- tabyl(sa_dhs_df2$hiv_test_ever_cat)
sa_dhs_sum <- sa_dhs_df2 |> group_by(hiv_test_ever) |>
  summarise(
    median_age = median(age, na.rm = T),
    mean_age = mean(age, na.rm = T),
    iqr_age = IQR(age, na.rm = T),
    mean_bmi = mean(bmi, na.rm = T),
    sd_bmi = sd(bmi, na.rm = T)
  )


```

```{r include = F}

sa_dhs_sum

sa_dhs_dem <- median(sa_dhs_sum$median_age)

sa_dhs_dem[1]

sa_descrip <- janitor::tabyl(sa_dhs_df2$hiv_test_ever)
sa_descrip
dim(sa_dhs_df2)

```


```{r include = F}

# By education levels

edu_df <- table(sa_dhs_df2$hiv_test_ever_cat,sa_dhs_df2$educ_level_cat)
edu_df
edu_df1 <- prop.table(edu_df, margin = 1)
edu_df1

# By residence type

res_df <- table(sa_dhs_df2$hiv_test_ever_cat,sa_dhs_df2$residence_type)
res_df
res_df1 <- prop.table(res_df, margin = 1)
res_df1

# By wealth index

wealth_df <- table(sa_dhs_df2$hiv_test_ever_cat,sa_dhs_df2$wealth_index_cat)
wealth_df
wealth_df1 <- prop.table(wealth_df, margin = 1)
wealth_df1

```
A total of `r nrow(sa_dhs_df2)` women were interviewed during the 2016 Demographic and Health Survey in South Africa. Of these, `r sa_descrip[2,2]` (`r round((sa_descrip[2,3]) * 100)`%) reported to have had HIV test before. 

## Education


Majority of the Women who had had an HIV test had secondary school education or higher compared to those who did not have an HIV test (**Figure 1** & **Figure 2**). Of the women who reported to have had an HIV test before, `r edu_df[1,3]`(`r round((edu_df1[1,3]) * 100)`%) had secondary school education. Similarly for those who never had an HIV test before, `r edu_df[2,3]`(`r round((edu_df1[2,3]) * 100)`%) had secondary school education.



```{r echo = F}


sa_dhs_df2 |> ggplot(mapping = aes(x  = hiv_test_ever_cat,
                                   fill = educ_level_cat)) +
  geom_bar(position = "fill") +
  labs(
    title = "Figure 1: HIV testing by education level",
    
    subtitle = "Relationship between  HIV testing and education level among women in South Africa",
    caption = "Source: 2016 South Africa Demographic and Health Survey(DHS)",
    y  = "Proportion",
    x = "Ever been tested for HIV",
    fill = "Education level"
  ) +
  theme_classic() +
  theme(plot.tag.position = "bottomright")
```


Except the category with no education, the median ages for those who tested for HIV were likely to be greater than the median ages for those who did not test for HIV across different education levels(**Figure 2**).


```{r echo = F}


# Figure 1: HIV testing by education level



sa_dhs_box_educ <- sa_dhs_df2 |> ggplot(mapping = aes(y = age,
                                                        x = hiv_test_ever_cat,
                                                        fill = hiv_test_ever_cat)) +
  geom_boxplot(alpha = 0.60,
               linewidth = 0.70) +
  theme_classic() +
  labs(
    title = "Figure 2:HIV testing associated with increased age in different education levels",
    subtitle = "Relationship between age and HIV testing by education level among women in South Africa",
    x = "HIV testing",
    y = " Age (in years)",
    caption = "Source: 2016 South Africa Demographic and Health Survey(DHS)",
    fill = "HIV Test Status"
  ) +
  scale_color_manual(values = c("#4C5B5C", "#6B8E7A", "#EFF2A9")) +
  facet_grid(~ educ_level_cat) +
  theme(axis.text.x = element_blank()) +
  geom_jitter(width = 0.2,
              alpha = 0.03) +
  stat_summary(
    fun = "mean",
    geom = "point",
    shape = 3,
    size = 1
  )

```

```{r echo = F}
sa_dhs_box_educ
```

## Residency



A total of `r res_df[2,1]`(`r round((res_df1[2,1]) * 100)`%) women who reported to have had an HIV test before resided in the rural areas while `r res_df[1,1]`(`r round((res_df1[1,1]) * 100)`%) women who reported not have had an HIV test resided in the rural areas(**Figure 3**).

```{r include = F}

# Figure 2: HIV testing by type of residency

sa_dhs_box_res <- sa_dhs_df2 |> ggplot(mapping = aes(x  = hiv_test_ever_cat,
                                   fill = residence_type)) +
  geom_bar(position = "fill") +
  labs(
    title = "Figure 3: HIV testing by type of residency",
    
    subtitle = "Relationship between  HIV testing and residency among women in South Africa",
    caption = "Source: 2016 South Africa Demographic and Health Survey(DHS)",
    y  = "Proportion",
    x = "Ever been tested for HIV",
    fill = "Residency type"
  ) +
  theme_classic() +
  theme(plot.tag.position = "bottomright")

```

```{r echo = F}
sa_dhs_box_res 
```

## Social economic status


Of the women who reported to have had an HIV test before, `r wealth_df[2,3]`(`r round((wealth_df1[2,3]) * 100)`%) were from middle-class households. Similarly for those who never had an HIV test before, `r wealth_df[1,3]`(`r round((wealth_df1[1,3]) * 100)`%) women were from middle-class households (**Figure 4**).



```{r echo = F}


sa_dhs_box_wealth <- sa_dhs_df2 |> ggplot(mapping = aes(
  y = age,
  x = reorder(hiv_test_ever_cat, age, median),
  fill = hiv_test_ever_cat
)) +
  geom_boxplot(alpha = 0.60,
               linewidth = 0.70) +
  theme_classic() +
  labs(
    title = "Figure 4: Increased HIV testing in different social economic strata",
    subtitle = "Relationship between age and HIV testing by wealth index among women in South Africa",
    x = "HIV testing",
    y = " Age (in years)",
    caption = "Source: 2016 South Africa Demographic and Health Survey(DHS)",
    fill = "HIV Test Status"
  ) +
  scale_color_manual(values = c("#4C5B5C", "#6B8E7A", "#EFF2A9")) +
  facet_grid(~ wealth_index_cat) +
  theme(axis.text.x = element_blank()) +
  geom_jitter(width = 0.2,
              alpha = 0.05) +
  stat_summary(
    fun = "mean",
    geom = "point",
    shape = 3,
    size = 1
  )

sa_dhs_box_wealth

```
```{r include = F}

sessionInfo()

```


# Conclusion

This capstone project aimed at evaluating the uptake of HIV testing by social demographic factors among women aged between 15 and 49 years in South Africa. The project findings include the following:

- HIV testing uptake increases with social economic status of the individuals e.g., education level and wealth index of the household

The recommendation for this project is to encourage interventions aimed at increasing levels of education among women. 

# Reference

- [The DHS Program: Demographic and Health Surveys](https://dhsprogram.com/)

